Estimating duration in the presence of misclassification

### 2024-08-01 cis-perfect-testing.tex
Christl p.91 comment

Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
\eta
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} d\lambda &\lambda \dist \GamDist\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

@@

If $i(j) = i_k$ then the event $O(j) = \vec{\nu}_k$ occurs if and only if the episode starts in the interval $[l^{(b)}_k, r^{(b)}_k]$ and ends in the interval $[l^{(e)}_k, r^{(e)}_k]$.
Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:
\begin{align}
p_{ik}
=& \prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)}, l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \right) \\
=& \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \times\prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} - b + 1 \leq D_{j} \leq r_k^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
\label{perf-test:eq:pia}
\end{align}
by the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

>>>>

Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
\eta
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} d\lambda &\lambda \dist \GamDist\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{substitute in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

@@

If $i(j) = i_k$ then the event $O(j) = \vec{\nu}_k$ occurs if and only if the episode starts in the interval $[l^{(b)}_k, r^{(b)}_k]$ and ends in the interval $[l^{(e)}_k, r^{(e)}_k]$.
Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:
\begin{align}
p_{ik}
=& \prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)}, l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \right) \\
=& \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \times\prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} - b + 1 \leq D_{j} \leq r_k^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by definition of $D_{j}$} \\
=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by definition of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
\label{perf-test:eq:pia}
\end{align}
by the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

### 2024-08-01 incidence-prevalence.tex
Christl p.91 comment

A similar argument will show that the correlation between $Y_t$ and $Y_{t'}$ ($t \neq t'$) is negligible.
The basic structure of this argument is as the previous one.
The result follows because the observation noise dominates, and this is uncorrelated.
In the following, all are implicitly conditioned on $\vec{Z}$.
However, for clarity I omit the conditioning from the notation.
\begin{align}
  &\lvert \cor(Y_t, Y_{t'}) \rvert \\
  &= \frac{\lvert\cov(Y_t, Y_{t'})\rvert}{\sqrt{\var(Y_t) \var(Y_{t'})}} &\text{by definition}\\
  &= \frac{\lvert \E(\cov(Y_t, Y_{t'} \mid \vec{P})) + \cov(\E(Y_t \mid \vec{P}), \E(Y_{t'} \mid \vec{P})) \rvert}{\sqrt{\var(Y_t) \var(Y_{t'})}} &\text{by law of total covariance} \\
  &= \frac{\lvert 0 + \cov(n_t P_t / \Npop, n_{t'} P_{t'} / \Npop) \rvert}{\sqrt{\var(Y_t) \var(Y_{t'})}} \\
  &= \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(Y_t) \var(Y_{t'})}}  \\
  &\ll \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(P_t) \var(P_{t'})}} &\text{as $\V(Y_t) \gg \V(P_t)$} \\
  &= \frac{n_t n_{t'}}{\Npop^2} \lvert \cor(P_t, P_{t'}) \rvert &\text{by def of $\cor$}\\
  &\leq \frac{n_t n_{t'}}{\Npop^2} &\text{as $\lvert \cor(\cdot) \rvert \leq 1$}.
\end{align}
$n_t n_{t'} / \Npop^2$ is very small, for the same reason that $n_t / \Npop$ is small.
$\lvert \cor(Y_t, Y_{t'}) \rvert$ is even smaller than this.
Therefore, neglecting the correlation between $Y_t$ and $Y_{t'}$ is a good approximation.

>>>>

A similar argument will show that the correlation between $Y_t$ and $Y_{t'}$ ($t \neq t'$) is negligible.
The basic structure of this argument is as the previous one.
The result follows because the observation noise dominates, and this is uncorrelated.
In the following, all are implicitly conditioned on $\vec{Z}$.
However, for clarity I omit the conditioning from the notation.
\begin{align}
  &\lvert \cor(Y_t, Y_{t'}) \rvert \\
  &= \frac{\lvert\cov(Y_t, Y_{t'})\rvert}{\sqrt{\var(Y_t) \var(Y_{t'})}} &\text{by definition}\\
  &= \frac{\lvert \E(\cov(Y_t, Y_{t'} \mid \vec{P})) + \cov(\E(Y_t \mid \vec{P}), \E(Y_{t'} \mid \vec{P})) \rvert}{\sqrt{\var(Y_t) \var(Y_{t'})}} &\text{by law of total covariance} \\
  &= \frac{\lvert 0 + \cov(n_t P_t / \Npop, n_{t'} P_{t'} / \Npop) \rvert}{\sqrt{\var(Y_t) \var(Y_{t'})}} \\
  &= \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(Y_t) \var(Y_{t'})}}  \\
  &\ll \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(P_t) \var(P_{t'})}} &\text{as $\V(Y_t) \gg \V(P_t)$} \\
  &= \frac{n_t n_{t'}}{\Npop^2} \lvert \cor(P_t, P_{t'}) \rvert &\text{by definition of $\cor$}\\
  &\leq \frac{n_t n_{t'}}{\Npop^2} &\text{as $\lvert \cor(\cdot) \rvert \leq 1$}.
\end{align}
$n_t n_{t'} / \Npop^2$ is very small, for the same reason that $n_t / \Npop$ is small.
$\lvert \cor(Y_t, Y_{t'}) \rvert$ is even smaller than this.
Therefore, neglecting the correlation between $Y_t$ and $Y_{t'}$ is a good approximation.
